### 2022-09-22 main/chapter_2ljets.tex
deduplicate atlas experiment citation; write trigger

Each trigger in \atlas' arsenal activates on a selection comprising some
combination of sufficiently hard objects.
The set of targeted objects includes all those used in this analysis and more:
electrons, muons, jets, $b$-jets, $\met$, as well as photons and hadronically
decaying tau leptons~\cite{atlas_PERF_2007_01}.

>>>>

Each trigger in \atlas' arsenal activates on a selection comprising some
combination of sufficiently hard objects.
The set of targeted objects includes all those used in this analysis and more:
electrons, muons, jets, $b$-jets, $\met$, as well as photons and hadronically
decaying tau leptons~\cite{atlas2008experiment}.

### 2022-09-22 main/chapter_context.tex
deduplicate atlas experiment citation; write trigger

\begin{quote}
``%
The ATLAS detector~\cite{PERF-2007-01} is a multipurpose particle detector with
a forward--backward symmetric cylindrical geometry and a near $4\pi$ coverage
in solid angle.%
\footnote{%
``ATLAS uses a right-handed coordinate system with its origin at the nominal
interaction point (IP) in the centre of the detector and the $z$-axis along the
beam pipe.
The $x$-axis points from the IP to the centre of the LHC ring, and the $y$-axis
points upwards.
Cylindrical coordinates $(r,\phi)$ are used in the transverse plane, $\phi$
being the azimuthal angle around the $z$-axis.
The pseudorapidity is defined in terms of the polar angle $\theta$ as
$\eta = -\ln \tan(\theta/2)$.
Angular distance is measured in units of
$\Delta R \equiv \sqrt{(\Delta\eta)^{2} + (\Delta\phi)^{2}}$.%
''\footnotemark~\cite{atlas2022searches}%
}
It consists of an inner tracking detector surrounded by a thin superconducting
solenoid providing a $2\,\mathrm{T}$ axial magnetic field, electromagnetic and
hadronic calorimeters, and a muon spectrometer.%
''~\cite{atlas2022searches}
\end{quote}
To illustrate this description, a vertical slice of \atlas\ is displayed in
Figure~\ref{fig:atlas_cutaway}, and each subsystem is described in more detail
in a subsection below:
the inner tracking detector in Section~\ref{sec:atlas_inner},
the electromagnetic and hadronic calorimeter systems in
Section~\ref{sec:atlas_calo},
and
the chambers of the muon spectrometer in Section~\ref{sec:atlas_muon}.
The final connection to the data analysed in this thesis is made through the
trigger system which is described in Section~\ref{sec:atlas_trigger},
and details of the data themselves are given in
Section~\ref{sec:atlas_data}

@@

Measurements of the tracks' curvature yield estimates of the muons' momenta;
although extremely energetic muons will have straight tracks,
\atlas' design and calibration achieves a high precision that increases to
only $10\%$ relative error on tracks with $1\,\eV[T]$ transverse momenta~\cite{
atlas1994proposal,
atlas2008experiment,
ATL-PHYS-PUB-2015-037
}.

\subsection{Trigger}
\label{sec:atlas_trigger}
\TODO{L1 HLT (L2, event thing)}
\TODO{trigger~\cite{atlas2016trigger}}
% 25ns bunch spacing
\TODO{
With $25\,\mathrm{ns}$ bunch spacing and \atlas' $12.5\,\textrm{m}$ radius,
light-speed particles ejected from one crossing are still zipping through
\atlas\ when the next collision arrives.
}
% fast decisions to not overflow buffer queue
\begin{quote}
``%
A two-level trigger system is used to select events.
The first-level trigger is implemented in hardware and uses a subset of the
detector information to accept events at a rate below $100\,\mathrm{kHz}$.
This is followed by a software-based trigger that reduces the accepted event
rate to $1\,\mathrm{kHz}$ on average depending on the data-taking conditions.%
''~\cite{atlas2022searches}
\end{quote}

>>>>

\begin{quote}
``%
The ATLAS detector~\cite{atlas2008experiment} is a multipurpose particle
detector with a forward--backward symmetric cylindrical geometry and a near
$4\pi$ coverage in solid angle.%
\footnote{%
``ATLAS uses a right-handed coordinate system with its origin at the nominal
interaction point (IP) in the centre of the detector and the $z$-axis along the
beam pipe.
The $x$-axis points from the IP to the centre of the LHC ring, and the $y$-axis
points upwards.
Cylindrical coordinates $(r,\phi)$ are used in the transverse plane, $\phi$
being the azimuthal angle around the $z$-axis.
The pseudorapidity is defined in terms of the polar angle $\theta$ as
$\eta = -\ln \tan(\theta/2)$.
Angular distance is measured in units of
$\Delta R \equiv \sqrt{(\Delta\eta)^{2} + (\Delta\phi)^{2}}$.%
''\footnotemark~\cite{atlas2022searches}%
}
It consists of an inner tracking detector surrounded by a thin superconducting
solenoid providing a $2\,\mathrm{T}$ axial magnetic field, electromagnetic and
hadronic calorimeters, and a muon spectrometer.%
''~\cite{atlas2022searches}
\end{quote}
To illustrate this description, a vertical slice of \atlas\ is displayed in
Figure~\ref{fig:atlas_cutaway}, and each subsystem is described in more detail
in a subsection below:
the inner tracking detector in Section~\ref{sec:atlas_inner},
the electromagnetic and hadronic calorimeter systems in
Section~\ref{sec:atlas_calo},
and
the chambers of the muon spectrometer in Section~\ref{sec:atlas_muon}.
The final connection to the data analysed in this thesis is made through the
trigger system which is described in Section~\ref{sec:atlas_trigger},
and details of the data themselves are given in
Section~\ref{sec:atlas_data}

@@

Measurements of the tracks' curvature yield estimates of the muons' momenta;
although extremely energetic muons will have straight tracks,
\atlas' construction and calibration achieves a high precision, which
increases to only $10\%$ relative error on tracks with $1\,\eV[T]$ transverse
momenta~\cite{
atlas1994proposal,
atlas2008experiment,
ATL-PHYS-PUB-2015-037
}.

\subsection{Trigger}
\label{sec:atlas_trigger}
The LHC's $25\,\mathrm{ns}$ bunch spacing is so fast that
light-speed particles ejected from one crossing are still zipping through
\atlas' $12.5\,\textrm{m}$ radius when the next collision arrives.
This extreme engineering is necessary to achieve the luminosities we
need to see rare effects, but to avoid becoming overwhelmed
by data it also demands a strict data reduction mechanism.
This mechanism is the trigger:
\begin{quote}
``%
A two-level trigger system is used to select events.
The first-level trigger is implemented in hardware and uses a subset of the
detector information to accept events at a rate below $100\,\mathrm{kHz}$.
This is followed by a software-based trigger that reduces the accepted event
rate to $1\,\mathrm{kHz}$ on average depending on the data-taking conditions.%
''~\cite{atlas2022searches}
\end{quote}
The first-level (L1) trigger uses its fast preliminary information to accept
interesting event and mark their regions of interest.
Complete data in these regions of interest are passed to the second-level
(high-level trigger, HLT) to inform its decision of whether to accept each
event.
Events accepted by the HLT are recorded for future analysis~\cite{
atlas2016trigger,
atlas2008experiment
}.
